### 2014-01-29 paper/paper.tex
more

\documentclass[a4paper,11pt,twocolumn]{article}

\usepackage[small,bf]{caption}
\usepackage{xltxtra}
\usepackage{times}

\title{North Sámi to South Sámi machine translation}
\author{Lene Antonsen \and Francis Tyers \and Trond Trosterud}
\date{}

\begin{document}

\maketitle

\section{Introduction}

\section{Languages}

% [Sámi language area map]

\section{Implementation}

% [Pipeline image]

\subsection{Analysis}

\subsubsection{Transfer}

\textbf{Bilingual dictionary}

\textbf{Rules}

\subsubsection{Generation}

\section{Evaluation}

\subsection{Statistics}

\subsection{Quantitative}

\subsection{Qualitative}

\subsection{User feedback}

\section{Future work}

\section{Conclusions}

\section*{Acknowledgements}

\section{References}

\end{document}

### 2014-01-30 paper/paper.tex
add some bib references
updates
more

@@
 \usepackage{xltxtra}
+\usepackage{fontspec}
+
+\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
+
 \usepackage{times}
+\usepackage{natbib}
 
-\title{North Sámi to South Sámi machine translation}
-\author{Lene Antonsen \and Francis Tyers \and Trond Trosterud}
+\setromanfont{Times New Roman}
+\setmonofont{FreeMono}
+
+\bibdata{paper}
+
+\title{A North Sámi to South Sámi machine translation prototype}
+\author{Lene Antonsen,\\Giellatekno,\\Tromsø\\{\tt lene.antonsen@uit.no}
+\and Francis Tyers,\\Giellatekno,\\Tromsø\\{\tt ftyers@dlsi.ua.es}
+\and Trond Trosterud,\\Giellatekno,\\Tromsø\\{\tt trond.trosterud@uit.no}}
 \date{}
@@
 \section{Introduction}
+
+% Previous work
+\cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
 
@@
 
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|r|}
+      \hline
+      Bilingual dictionary ({\tt sme}$\rightarrow${\tt sma}) & 15,204 \\ % 10776 proper names
+      Transfer rules ({\tt sme}$\rightarrow${\tt sma}) & 57 \\
+      \hline
+    \end{tabular}
+    \label{table:transfer}
+    \caption{Number of bilingual dictionary entries and transfer rules}
+  \end{center}
+\end{table}
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|r|}
+      \hline
+      \textbf{Corpus} & \textbf{Coverage}  \\
+      \hline
+      Schoolbooks     & 0.0 $\pm$ 0.0 \\
+      General         & 0.0 $\pm$ 0.0 \\
+      \hline
+    \end{tabular}
+    \label{table:coverage}
+    \caption{Vocabulary coverage of the system}
+  \end{center}
+\end{table}
+
 \subsection{Quantitative}
@@
 
-\section{References}
+\bibliographystyle{apalike}
+\bibliography{paper}
 

### 2014-02-05 paper/paper.tex
Some possible points.

@@
 \section{Introduction}
+Suggestions from Lene:  
+\begin{enumerate}
+\item main point: 
+\begin{itemize}
+\item translation nob-sma which actually is nob-sme-sma (should somehow be in the title?) 
+\item the evaluators evaluate nob-sma
+\end{itemize}
+\item subpoints: 
+\begin{itemize}
+\item larger syntactic differences between SL an TL than usually done in Apertium?
+\item little sma lexical resources, we have to find words by comparing translated texts
+\item reaction from sma linguist in the normgiving organ: she sees the possibility of using MT as a help for discussing terminology and getting the best ones into use
+\item the language socity is used to the impact from the mojority languages, a new controversial(?) thought is the linguistic impact from another minority lanugage 
+\end{itemize}
+\end{enumerate}
 

### 2014-02-07 paper/paper.tex
More
meeting

@@
 \setromanfont{Times New Roman}
-\setmonofont{FreeMono}
+%\setmonofont{FreeMono}
 
@@
 \section{Introduction}
-Suggestions from Lene:  
-\begin{enumerate}
-\item main point: 
-\begin{itemize}
-\item translation nob-sma which actually is nob-sme-sma (should somehow be in the title?) 
-\item the evaluators evaluate nob-sma
-\end{itemize}
-\item subpoints: 
-\begin{itemize}
-\item larger syntactic differences between SL an TL than usually done in Apertium?
-\item little sma lexical resources, we have to find words by comparing translated texts
-\item reaction from sma linguist in the normgiving organ: she sees the possibility of using MT as a help for discussing terminology and getting the best ones into use
-\item the language socity is used to the impact from the mojority languages, a new controversial(?) thought is the linguistic impact from another minority lanugage 
-\end{itemize}
-\end{enumerate}
+%@trond
 
@@
 \cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
+% LREC?
+
+%pivot translation references --babych paper
 
 \section{Languages}
-
+%@trond
 % [Sámi language area map]
@@
 \section{Implementation}
-
+%@fran
 % [Pipeline image]
@@
 
+The bilingual dictionary
+general dictionary
+general words
+special domain
+
+\item problems caused by this model: ex from two terms to one term to two terms
+
+Example: nob-1,2   sme-x   sma-1,2
+
+Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts
+
 \textbf{Rules}
+
+\item larger syntactic differences between SL an TL than usually done in Apertium?
 
@@
 \section{Evaluation}
+%@ lene,fran
+\item translation nob-sma which actually is nob-sme-sma (should somehow be in the title?) 
+\item the evaluators evaluate nob-sma
 
 \subsection{Statistics}
-
+%@fran
 \begin{table}
@@
 \subsection{Quantitative}
-
+%@lene
 \subsection{Qualitative}
+%@lene
 
 \subsection{User feedback}
+%@lene
 
 \section{Future work}
-
+%@trond
 \section{Conclusions}
+%@trond
+\item reaction from sma linguist in the normgiving organ: she sees the possibility of using MT as a help for discussing terminology and getting the best ones into use
+\item the language socity is used to the impact from the mojority languages, a new controversial(?) thought is the linguistic impact from another minority lanugage
 

### 2014-02-07 paper/paper.tex
now evening.

@@
 %@trond
+%To be written after the paper is finished, but try to state the idea.
+
+The paper presents a machine translation system from North Saami to
+South Saami. The system is intended to work in a translation setting
+where North Saami acts as a pivot language, with manual translation
+from Norwegian into the largest of the Saami languages, and thereafter
+to offer machine translation service, with postediting, to the other,
+smaller, Saami languages. On the one hand, the Saami languages are
+closely related, and therefore lend themselves better to MT than a
+system translating from Norwegian directly, but on the other hand, the
+classical problems of translating via a pivot language apply in this
+case as well.
+
+The paper is structured as follows. Afther looking at previous work
+and at the languages themselves, we give a presentation of the actual
+machine translation system. Thereafter comes an evaluation of the
+system. We presented translated text to translators, alongside with
+the Norwegian original, without revealing the fact that the texts were
+actually translated from North Saami.
+
+finally comes a conclusion.
 
@@
 % [Sámi language area map]
+
+Norwegian North Germanic. The Saami languages constitute the
+westernmost branch of the Uralic language family. They contain most of
+the classical Uralic characteristica: A rich verbal morphology (3
+numbers and persons, a rich repertoire of infinite forms), and a
+medium-size case system with both grammatical and adverbial functions,
+and no gender distinction. They also show extensive contact with their
+Germanic neighbours. Many grammatical structures ahre head-initial
+constructions, as compared to the classical Uralic pattern, and the
+tense system is compatible with the North Germanic one.
+
+There are also differences within the Saami family.
+
+sme-sma.
 

### 2014-02-08 paper/paper.tex
Added some examples to Fran's part

@@
 and no gender distinction. They also show extensive contact with their
-Germanic neighbours. Many grammatical structures ahre head-initial
+Germanic neighbours. Many grammatical structures are head-initial
 constructions, as compared to the classical Uralic pattern, and the
@@
 
-\item problems caused by this model: ex from two terms to one term to two terms
+\subsubsection{Problems caused by this model: ex from two terms to one term to two terms}
 
 Example: nob-1,2   sme-x   sma-1,2
+example the verbs ‘read, count, say’, nob: \textit{lese, telle, uttale}, sme: \textit{lohkat} can be used about all three, sma: \textit{lohkedh, ryöknehtidh, jiehtedh}
 
-Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts
+Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts (nob-sme with nob-sma)
+
+\subsubsection{Problems combining sma-nob and sme-nob dictionary}
+This was done, it had to be manually edited because of nob-words with more than one meaning. 
+Example: nob \textit{regne} means both 'to rain' and ‘to calculate’, 
+which are two different verbs both in sme \textit{arvit, rehkenastit} and in sma \textit{abrodh, ryöknedh}
 
@@
 
-\item larger syntactic differences between SL an TL than usually done in Apertium?
+larger syntactic differences between SL an TL than usually done in Apertium? \\
+This also has consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
+ex: nob ‘å trekke lodd’ sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
+nob ‘å presentere’ sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
+I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
+fremdeles må forklare innholdet.
 

### 2014-02-08 paper/paper.tex
ref

@@
 Norwegian North Germanic. The Saami languages constitute the
-westernmost branch of the Uralic language family. They contain most of
+westernmost branch of the Uralic language family. They possess most of
 the classical Uralic characteristica: A rich verbal morphology (3
@@
 sme-sma.
+
+An overview \cite{sammallahti98}.
 
@@
 %@ lene,fran
-\item translation nob-sma which actually is nob-sme-sma (should somehow be in the title?) 
-\item the evaluators evaluate nob-sma
+ translation nob-sma which actually is nob-sme-sma (should somehow be in the title?) 
+ the evaluators evaluate nob-sma
 
@@
 %@trond
-\item reaction from sma linguist in the normgiving organ: she sees the possibility of using MT as a help for discussing terminology and getting the best ones into use
-\item the language socity is used to the impact from the mojority languages, a new controversial(?) thought is the linguistic impact from another minority lanugage
+
+ reaction from sma linguist in the normgiving organ: she sees the possibility of using MT as a help for discussing terminology and getting the best ones into use
+
+ the language socity is used to the impact from the mojority languages, a new controversial(?) thought is the linguistic impact from another minority lanugage
 

### 2014-02-08 paper/paper.tex
A bit more about the languages.

@@
 
-There are also differences within the Saami family.
+North and South Saami are not mutually intellegible, due both to
+linguistic distance and to radically different prinicples for their
+literary languages. An analogy would be the difference between English 
+and Frisian.
 
-sme-sma.
+Grammatical similarities between North and South Saami include the 
+following: Both languages have 3 persons and numbers, they have a similar
+system for postpositions and verb derivation. Noun phrase syntax is 
+similar, and the case systems are almost identical.
 
-An overview \cite{sammallahti98}.
+Taking a closer look, there are still differences. In both North and
+South Saami, negation is expressed with a negation verb which inflects
+for person and number, but in 
+South Saami, this verb is inflected for Tense as well. South Saami
+has OV word order and lacks a copula in predicative constructions, 
+whereas North Saami uses VO in simple sentences, and requires a copula.
+
+Although the case systems are similar, North Saami has a locative case
+that covers the semantic field of both inessive and elative in 
+South Saami, the locative must thus be split into two cases. There is
+also some differences in case usage, plural objects are accusative
+in North Saami, but nominative for indefinite and accusative for 
+definite objects in South Saami.
+
+For a machine translation system the ortographic differences 
+imply that apart from person
+names and acronyms there will be no free rides during the conversion 
+process. The core vocabulary is distinct, even recent loans from
+the same donor languages are different, and the vocabulary coverage
+for a working system must thus be very good.
+
+An overview over the differences between the two languages can
+be found in \cite{sammallahti98}.
 

### 2014-02-09 paper/paper.tex
add cov
more

@@
 
-For a machine translation system the ortographic differences 
+For a machine translation system the orthographic differences 
 imply that apart from person
@@
 
+The translator was implemented using the Apertium platform \cite{forcada2011}. Apertium 
+provides a highly modular set of tools for building rule-based
+machine translation systems. Apertium language pairs are set up as Unix pipelines, where the
+typical pipeline consists of:
+
+\begin{itemize}
+\item deformatting (encapsulating formatting/markup from the engine),
+\item source-language (SL) morphological analysis with a finite-state
+  transducer (FST),
+\item disambiguation using a Hidden Markov Model (HMM) and/or
+  Constraint Grammar (CG), 
+\item lexical transfer (word-translation on the disambiguated source),
+\item lexical selection (choosing the appropriate word out of a set of possible
+   translations),
+\item one or more levels of finite-state based structural transfer
+  (reordering, and changes to morphological features),
+\item target-language (TL) generation with an FST
+\item reformatting (unencapsulating format information)
+\end{itemize}
+
+See Figure~\ref{fig:pipeline} for an overview of the modules used
+in this particular language pair.
+
+\begin{figure*}
+ \centering
+ \includegraphics[width=0.75\textwidth]{architecture-overview.pdf}
+ \caption{Overview of the translation pipeline.}
+ \label{fig:pipeline}
+\end{figure*}
+
 \subsection{Analysis}
 
-\subsubsection{Transfer}
+\subsection{Transfer}
 
 \textbf{Bilingual dictionary}
+
+Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
+ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
+nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
+I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
+fremdeles må forklare innholdet.
 
@@
 
-\textbf{Rules}
+\subsubsection{Rules}
 
-larger syntactic differences between SL an TL than usually done in Apertium? \\
-This also has consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
-ex: nob ‘å trekke lodd’ sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
-nob ‘å presentere’ sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
-I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
-fremdeles må forklare innholdet.
+The syntactic differences between North Saami and South Saami are
+greater than are usually dealt with between related language pairs
+in Apertium. 
+In order to be able to transfer VO structures to OV more reliably,
+the transfer phase is split into five parts instead of the three
+more typically used in Apertium:
+
+%t1x local chunking NP, VP
+%t2x localcoordination NP+NP, etc.
+%t3x empty, but for adpositional phrases + Rel clauses
+%t4x Constituent mvm, SVO -> SOV
+%t5x Cleanup
+
+\begin{itemize} 
+  \item \texttt{chunker}: Chunk input words into groups, e.g. noun groups, verb chains
+  \item \texttt{interchunk1}: Merge chunks which have local coordination, e.g. the sequence [NP $x$] [CC and] [NP $y$] is merged into [NP $x$ and $y$]
+  \item \texttt{interchunk2}: Merges relative clauses and adpositional phrases.
+  \item \texttt{interchunk3}: Reorder constituents, e.g. SVO $\rightarrow$ SOV.
+  \item \texttt{postchunk}: Cleanup
+\end{itemize}
 
@@
   \begin{center}
-    \begin{tabular}{|l|r|}
+    \begin{tabular}{|l|r|r|}
       \hline
-      \textbf{Corpus} & \textbf{Coverage}  \\
+      \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage (\%)}  \\
       \hline
-      Schoolbooks     & 0.0 $\pm$ 0.0 \\
-      General         & 0.0 $\pm$ 0.0 \\
+      Schoolbooks     & 468,697 & 92.4 $\pm$ 0.58 \\
+      General         & & 0.0 $\pm$ 0.0 \\
       \hline

### 2014-02-09 paper/paper.tex
more

@@
 
+Morphological analysis is done on the input using the Helsinki Finite-State
+Toolkit \cite{linden2011}. For each surface form, a finite-state transducer
+returns a set of the possible analyses, where an analysis is a combination
+of lemma, and a sequence of tags which describe the morphological structure
+of the surface form.
+
+A Constraint Grammar-based disambiguator then selects the most appropriate 
+analysis for each surface form according to the context, and assigns to each 
+analysis a syntactic tag denoting its syntactic function (subject, object, 
+main verb, \ldots).
+
 \subsection{Transfer}
 
-\textbf{Bilingual dictionary}
+\subsubsection{Lexical transfer}
 
@@
 
-\subsubsection{Problems caused by this model: ex from two terms to one term to two terms}
+\textbf{Problems caused by this model: ex from two terms to one term to two terms}
 
@@
 
-\subsubsection{Problems combining sma-nob and sme-nob dictionary}
+\textbf{Problems combining sma-nob and sme-nob dictionary}
 This was done, it had to be manually edited because of nob-words with more than one meaning. 
@@
 
-\subsubsection{Rules}
+\subsubsection{Lexical selection}
+
+\subsubsection{Structural transfer}
 
@@
 \subsubsection{Generation}
+
+Generation is done using a finite-state transducer, also 
+compiled using HFST. Each lexical unit which is output by
+the postchunk module is looked up in the generation transducer
+and the surface form is output.
 
@@
 
+Linda Wiechetek, Divvun
+
 \bibliographystyle{apalike}

### 2014-02-09 paper/paper.tex
e
fonts
add abstract
updates
Meir tekst.

@@
 
-\usepackage[small,bf]{caption}
 \usepackage{xltxtra}
@@
 
+\usepackage[small,bf]{caption}
+
+\setmainfont[BoldFont={Times New Roman}]{Times New Roman}
 \setromanfont{Times New Roman}
@@
 \maketitle
+
+\begin{abstract}
+This paper describes the development of a rule-based
+machine translation system from North Sámi to South
+Sámi, and its evaluation in a setting where North Sámi
+functions as pivot translation via manual translation
+from Norwegian North Sámi, and thereafter MT to South Sámi. 
+\end{abstract}
 
@@
 %pivot translation references --babych paper
+\cite{babych2007}
 
@@
 
-Norwegian North Germanic. The Saami languages constitute the
-westernmost branch of the Uralic language family. They possess most of
-the classical Uralic characteristica: A rich verbal morphology (3
-numbers and persons, a rich repertoire of infinite forms), and a
-medium-size case system with both grammatical and adverbial functions,
-and no gender distinction. They also show extensive contact with their
-Germanic neighbours. Many grammatical structures are head-initial
+As a Germanic language, Norwegian differs from the Saami languages
+in numerous respects. It has definiteness as a morphological category,
+no case, prepositions and verb / particle constructions, and a 
+relatively strict word order, with V2 in main clauses. Norwegian
+and the Saami languages also show Sprachbund phenomena, though.
+Most notably, they have the same tense system. Whereas the Saami
+languages also possess a number of infinite verb construction, the
+Norwegian embedded clause pattern is in most cases a possible option
+for the Saami languages as well.
+
+The Saami languages constitute the westernmost branch of the Uralic
+language family. They possess most of the classical Uralic
+characteristica: A rich verbal morphology (3 numbers and persons, a
+rich repertoire of infinite forms), and a medium-size case system with
+both grammatical and adverbial functions, and no gender
+distinction. They also show extensive contact with their Germanic
+neighbours. Many grammatical structures are head-initial
 constructions, as compared to the classical Uralic pattern, and the
@@
 Grammatical similarities between North and South Saami include the
-following: Both languages have 3 persons and numbers, they have a similar
-system for postpositions and verb derivation. Noun phrase syntax is 
-similar, and the case systems are almost identical.
+following: Both languages have 3 persons and numbers, they have a
+similar system for postpositions and verb derivation. Noun phrase
+syntax is similar, and the case systems are almost identical.
 
@@
 South Saami, negation is expressed with a negation verb which inflects
-for person and number, but in 
-South Saami, this verb is inflected for Tense as well. South Saami
-has OV word order and lacks a copula in predicative constructions, 
-whereas North Saami uses VO in simple sentences, and requires a copula.
+for person and number, but in South Saami, this verb is inflected for
+Tense as well. South Saami has OV word order and lacks a copula in
+predicative constructions, whereas North Saami uses VO in simple
+sentences, and requires a copula.  The possession construction is
+different from the Norwegian one, normally using copula rather than
+possessor subject + a verb \textit{to have}. In North Saami, the
+construction is 
+\textit{locative possessor + copula + possessed in nominative}, 
+whereas the possessor is in the genitive in South
+Saami.
 
@@
 
-For a machine translation system the orthographic differences 
-imply that apart from person
-names and acronyms there will be no free rides during the conversion 
-process. The core vocabulary is distinct, even recent loans from
-the same donor languages are different, and the vocabulary coverage
-for a working system must thus be very good.
+For a machine translation system the orthographic differences imply
+that apart from person names and acronyms there will be no free rides
+during the conversion process. The core vocabulary is distinct, even
+recent loans from the same donor languages are different, and the
+vocabulary coverage for a working system must thus be very good.
 
@@
 %@fran
-% [Pipeline image]
 
@@
 \subsubsection{Lexical transfer}
-
+%@lene
 Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.

### 2014-02-09 paper/paper.tex
Litt mer

@@
 %@lene
+
+Some numbers -- kladd
+\begin{itemize}
+\item 3568 general word pairs + 61 proper nouns
+\item added: 873 special domain word pairs 
+\item added: 10 721 proper nouns
+\end{itemize}
+
+There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a North Saami-Norwegian dictionary and a South Saami-Norwegian dictionary.
+The word pairs were manually edited and the work revealed many incorrect pairs because of nob-words with more than one meaning.
+Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}.
+The result of this work was 3568 general word pairs and 61 proper nouns.
+
 Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
@@
 nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
-I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
-fremdeles må forklare innholdet.
+I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma fremdeles må forklare innholdet ved hjelp av flere ord.
 
-The bilingual dictionary
-general dictionary
-general words
+The bilingual dictionary \\
+general dictionary\\
+general words\\
 special domain
@@
 Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts (nob-sme with nob-sma)
-
-\textbf{Problems combining sma-nob and sme-nob dictionary}
-This was done, it had to be manually edited because of nob-words with more than one meaning. 
-Example: nob \textit{regne} means both 'to rain' and ‘to calculate’, 
-which are two different verbs both in sme \textit{arvit, rehkenastit} and in sma \textit{abrodh, ryöknedh}
 

### 2014-02-09 paper/paper.tex
La til LREC paper, for tall for  sme-analysator
Litt mer

@@
 \cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
-% LREC?
+% LREC for numbers for the sme-analyser:
+\cite{AntonsenEtalReusing2010}
 
@@
 
-Some numbers -- kladd
+The bilingual dictionary: Some numbers 
 \begin{itemize}
@@
 
-There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a North Saami-Norwegian dictionary and a South Saami-Norwegian dictionary.
-The word pairs were manually edited and the work revealed many incorrect pairs because of nob-words with more than one meaning.
+There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a general North Saami-Norwegian dictionary and a general South Saami-Norwegian dictionary.
+The word pairs were manually edited and the work revealed many incorrect pairs because of Norwegian words with more than one meaning.
 Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}.
@@
 
-Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
-ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
-nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
-I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma fremdeles må forklare innholdet ved hjelp av flere ord.
+The existing Norwegian-South Saami lexical resources are poor, with little terminology from the modern society.
+South Saami has been used more for giving information to the Saami people especially the last two years, and there are some texts especially about school administration and curriculum on Internet.
+By comparing Norwegian texts with the translations to North Saami and South Saami, we added 873 special domain word pairs to the bilingual dictionary.
 
-The bilingual dictionary \\
-general dictionary\\
-general words\\
-special domain
+Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami object+verb og adverb+verb.
+Examples are Norwegian `å trekke lodd' is translated into North Saami \textit{vuorbádit} and North Saami \textit{vuerpiem giesedh}.
 
-\textbf{Problems caused by this model: ex from two terms to one term to two terms}
+Norwegian `å presentere' is in North Saami \textit{ovdanbuktit} but in South Saami \textit{åvtese buektedh}.
+Often is the reason to this difference that one have utviklet North Saami terms to match the Norwegian terms, and in South Saami one still has to explain the meaning of the term with more words.
 
-Example: nob-1,2   sme-x   sma-1,2
-example the verbs ‘read, count, say’, nob: \textit{lese, telle, uttale}, sme: \textit{lohkat} can be used about all three, sma: \textit{lohkedh, ryöknehtidh, jiehtedh}
+These are solved by adding giving the translation as MWE to the morpological analyser.
 
-Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts (nob-sme with nob-sma)
+The pivot model causes some lexical problems.
+Even if the lexical inndeling often is similar in South Saami and North Saami, as in for the ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
+But in South Saami is no common verb for these three, and the lexical xxx is like in Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
 
 \subsubsection{Lexical selection}
+Because we were tilpasse the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
 

### 2014-02-09 paper/paper.tex
xxx

@@
 The bilingual dictionary: Some numbers
+
 \begin{itemize}
@@
 The existing Norwegian-South Saami lexical resources are poor, with little terminology from the modern society.
-South Saami has been used more for giving information to the Saami people especially the last two years, and there are some texts especially about school administration and curriculum on Internet.
+South Saami has been used more for giving information to the Saami people especially during the last two years, and there are some texts especially about school administration and curriculum on Internet.
 By comparing Norwegian texts with the translations to North Saami and South Saami, we added 873 special domain word pairs to the bilingual dictionary.
 
-Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami object+verb og adverb+verb.
-Examples are Norwegian `å trekke lodd' is translated into North Saami \textit{vuorbádit} and North Saami \textit{vuerpiem giesedh}.
+Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami as $object+verb$ and $adverb+verb$ strings.
+Examples include Norwegian `å trekke lodd', («draw lots for») which is translated into North Saami as one verb \textit{vuorbádit}, but into South Saami as object+verb \textit{vuerpiem giesedh} (lit.
+«lot draw»).
 
@@
 The pivot model causes some lexical problems.
-Even if the lexical inndeling often is similar in South Saami and North Saami, as in for the ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
-But in South Saami is no common verb for these three, and the lexical xxx is like in Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
+Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
+But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
 
 \subsubsection{Lexical selection}
-Because we were tilpasse the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
+Because we were adapting the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
 

### 2014-02-09 paper/paper.tex
La til WER

@@
 
-The pivot model causes some lexical problems.
+\subsubsection{Lexical selection}
+Because we were adapting the MT system to a special domain, we only made xxx rules for lexical selection. An MT-system for more domains, will certainly need more such rules.
+
+The pivot model causes some extra lexical selection.
 Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
 But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
-
-\subsubsection{Lexical selection}
-Because we were adapting the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
 
@@
 \subsection{Quantitative}
-%@lene
+
+WER -- word error rate
+\begin{table}
+  \begin{center}
+    \begin{tabular}{|l|r|r|r|}
+      \hline
+      \textbf{ } & \textbf{Eval. A} & \textbf{Eval. B} & \textbf{Eval. C}  \\
+      \hline
+word order in account & 62,37  &  49,46  &  52,69 \\
+word order not in account  &  36,20   &  28,32  &  28,32 \\
+      \hline
+    \end{tabular}
+    \label{table:coverage}
+    \caption{Vocabulary coverage of the system}
+  \end{center}
+\end{table}
+
 \subsection{Qualitative}

### 2014-02-09 paper/paper.tex
Added numbers from the oovtast presentations.
updates
update

@@
 \begin{itemize}
-\item 3568 general word pairs + 61 proper nouns
+\item 3,568 general word pairs + 61 proper nouns
 \item added: 873 special domain word pairs 
@@
 Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}.
-The result of this work was 3568 general word pairs and 61 proper nouns.
+The result of this work was 3,568 general word pairs and 61 proper nouns.
 
@@
 The pivot model causes some extra lexical selection.
-Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
+Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calculate’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
 But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
@@
       \hline
-      \textbf{ } & \textbf{Eval. A} & \textbf{Eval. B} & \textbf{Eval. C}  \\
+      & \textbf{Eval. A} & \textbf{Eval. B} & \textbf{Eval. C}  \\
       \hline
-word order in account & 62,37  &  49,46  &  52,69 \\
-word order not in account  &  36,20   &  28,32  &  28,32 \\
+\textbf{WER} (\%) & 62.37  &  49.46  &  52.69 \\
+\textbf{PER} (\%) &  36.20   &  28.32  &  28.32 \\
       \hline
@@
     \label{table:coverage}
-    \caption{Vocabulary coverage of the system}
+    \caption{Evaluation metrics.}
   \end{center}
@@
 %@lene
+
+From the oovtast-presentation:
+\begin{table*} 
+\begin{center}
+\begin{tabular}{|l|r||r|r|r|}
+\hline
+\textbf{ }  & \textbf{MT} &  \textbf{A} & \textbf{B} & \textbf{C} \\
+\hline
+Lexical differences: & &   37 & 27 & 19 \\
+\hline
+Morphological differences: & &   15 & 14 & 16 \\
+\hline
+\hline
+\multicolumn{5}{|l|} {\textbf{Syntax} }   \\
+\hline
+SOV/SAdvlV & 4 &  10 & 12 & 12 \\
+\hline
+Det pronomen & 4 &   7 & 4 & 7 \\
+\hline
+Ind. articles & 0 &   7 & 0 & 1 \\
+\hline
+Complement instead of genitive mod. & -- &   2 & 2 & 0 \\
+\hline
+Habitive \textit{utnedh} & 0 &  5 & 2 & 2 \\
+\hline
+\hline
+All together & 8 &   31 & 20 & 21 \\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
 

### 2014-02-09 paper/paper.tex
update

@@
 
+We evaluate the system for both vocabulary coverage, and translation quality.
+For vocabulary coverage, we take two corpora, one in domain (\texttt{schoolbooks}) and 
+one general domain.
+
+For translation quality, we selected an text in the domain of school books. The 
+text consisted of 300 words. This was then translated by the system and the resulting
+translation, and the Norwegian original was given to three translators. They were
+ asked to postedit it to make an adequate South Sámi translation, and also to fill
+in a feedback questionnaire.
+
 \subsection{Statistics}

### 2014-02-09 paper/paper.tex
update
About the sme analyser, and two possible refs for sma, on for the socioling situation. Make a motivation section?

@@
 \cite{babych2007}
+
+\section{Motivation}
+sma sociolinguistic situation \cite{Todalsvahken2007}
 
@@
 
+The North Saami analyser has an F-score of 0.93 for morphological disambiguation
+and assignment fof grammatical functions \cite{AntonsenEtalReusing2010}.
+
 \subsection{Transfer}
@@
 
+Here is some facts about the South Saami analyser, if we need it \cite{smadis2011}.
+
 \section{Evaluation}
@@
 %@lene
+\subsubsection{Lexicon}
+The Norwegian text consisted of 288 words. The three evaluators had chosen respectively 42, 30 and 23 other words than the mt-system, for only for 9 words all the evaluators had chosen another word.
+
+\subsubsection{Morphology}
+
+\subsubsection{Syntax}
 
 From the oovtast-presentation:
-\begin{table*} 
+\begin{table*}[h] 
 \begin{center}
@@
 \hline
-Lexical differences: & &   37 & 27 & 19 \\
+Lexical differences: & &   42 & 30 & 23 \\
 \hline
-Morphological differences: & &   15 & 14 & 16 \\
+Morphological differences: & &   7 & 7 & 6 \\
 \hline
@@
 \hline
-All together & 8 &   31 & 20 & 21 \\
-\hline
 \end{tabular}
 \end{center}
+\label{table:qualitative}
+\caption{A break down of translation errors.}
 \end{table*}

### 2014-02-10 paper/paper.tex
Debugging bibtex, but still not there. Should the example package be Covington? Probbly not.

@@
 \usepackage[small,bf]{caption}
+
+\usepackage{covington}
 
@@
 \begin{abstract}
+
 This paper describes the development of a rule-based
@@
 from Norwegian North Sámi, and thereafter MT to South Sámi.
+
 \end{abstract}
@@
 
-% Previous work
-\cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
-% LREC for numbers for the sme-analyser:
+Previous work
+
+\cite{tyers09} and \cite{wiechetek10} and \cite{trosterud12}
+
+LREC for numbers for the sme-analyser:
+
 \cite{AntonsenEtalReusing2010}
 
-%pivot translation references --babych paper
+pivot translation references --babych paper
+
 \cite{babych2007}
@@
 \section{Motivation}
+
 sma sociolinguistic situation \cite{Todalsvahken2007}
@@
 The word pairs were manually edited and the work revealed many incorrect pairs because of Norwegian words with more than one meaning.
-Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}.
+Example is the Norwegian verb `regne', which means both `to rain' and `to calculate', which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}.
 The result of this work was 3,568 general word pairs and 61 proper nouns.
@@
 The pivot model causes some extra lexical selection.
-Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calculate’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}.
+Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial `to rain, to calculate' pair, there are also counter examples, like the Norwegian verbs `lese, telle, uttale' («read, count, utter») which all three can be expressed with the North Saami verb \textit{lohkat}.
 But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.
@@
     \end{tabular}
-    \label{table:coverage}
+    \label{table:coverage2}
     \caption{Evaluation metrics.}
@@
 %@lene
-\subsubsection{Lexicon}
-The Norwegian text consisted of 288 words. The three evaluators had chosen respectively 42, 30 and 23 other words than the mt-system, for only for 9 words all the evaluators had chosen another word.
+\subsubsection{Lexicon and morphology}
+The Norwegian text consisted of 288 words. The three evaluators had chosen respectively 42, 30 and 23 other words than the MT-system, for only for 9 words all the evaluators had chosen another word.
 
-\subsubsection{Morphology}
+The morphology is quite good, but in four cases an incorrect or lack of disambiguation of the North Saami text has resultet in wrong morphological form to be generated.
+Some North Saami verbs require a certain case, which is not the right one for the corresponding South Saami verb.
 
 \subsubsection{Syntax}
+The evaluation revealed that the syntax was the poorest part of the MT-translation.
+
+SOV: The evaluators used 10--12 SOV or SAdvlV constructions in there texts, in the MT-text there were only 4, because of ....
+
+The MT-system generated only the construction \textit{genitive possessor + copula + possessed in nominative} for possessive, but the evaluators preferred to some extent also using the verb \text{nominative possessor + utnedh + possessed in accusative}, one of the evaluators five times, the others twice.
+The MT-text didn't generate indefinite articles, but one of the evaluators used it six times, another one used it once and the last one didn't use it at all.
+In all cases there was indefinite article in the Norwegian source text.
+
+In North Saami there is no defiteness expressed in the morphology of the noun.
+But in South Saami it can be expressed in the object by choosing plural nominative instead of plural accusative.
+This is something which was not made rules for in the MT-system, and in one such case all the evaluators disagree with the MT-text.
+
+A complement in Norwegian, like ‘tilbud fra en lærebedrift’ is usually in North Saami tranlated to a genitive modifyer \textit{oahppofitnodaga+Gen fálaldaga+Acc}, but in South Saami translations it is quite usual to translate it with a complement \textit{faalenassem+Acc aktede+Art learoesïelteste+Ela}, but we see from the translations that also the North Saami construction is possible: learoesïelten faalenassem
+
+\ex.
+\a. Norwegian: tilbud (Obj) fra en lærebedrift ‘offer from a xxxxcompany’\\
+\a. North Saami: \textit{oahppofitnodaga+Gen fálaldaga+Acc}\\
+\a. South Saami 1: \textit{faalenassem+Acc aktede+Art learoesïelteste+Ela}
+\a. South Saami 2: \textit{learoesïelten+Gen faalenassem+Acc}
 
@@
 \hline
-Lexical differences: & &   42 & 30 & 23 \\
+Lexical differences: & &   43 & 30 & 23 \\
 \hline
@@
 \hline
-Ind. articles & 0 &   7 & 0 & 1 \\
+Ind. articles & 0 &   6 & 0 & 1 \\
 \hline

### 2014-02-10 paper/paper.tex
Reverting to last version , it seems Fran and I have different xetex.

@@
 \usepackage[small,bf]{caption}
-
-\usepackage{covington}
 
@@
 \begin{abstract}
-
 This paper describes the development of a rule-based
@@
 from Norwegian North Sámi, and thereafter MT to South Sámi. 
-
 \end{abstract}
@@
 
-Previous work
-
-\cite{tyers09} and \cite{wiechetek10} and \cite{trosterud12}
-
-LREC for numbers for the sme-analyser:
-
+% Previous work
+\cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
+% LREC for numbers for the sme-analyser:
 \cite{AntonsenEtalReusing2010}
 
-pivot translation references --babych paper
-
+%pivot translation references --babych paper
 \cite{babych2007}
@@
 \section{Motivation}
-
-sma sociolinguistic situation \cite{Todalsvahken2007}
+sma sociolinguistic situation \citep{Todalsvahken2007}
 
@@
 The North Saami analyser has an F-score of 0.93 for morphological disambiguation
-and assignment fof grammatical functions \cite{AntonsenEtalReusing2010}.
+and assignment fof grammatical functions \citep{AntonsenEtalReusing2010}.
 
@@
 
-Here is some facts about the South Saami analyser, if we need it \cite{smadis2011}.
+Here is some facts about the South Saami analyser, if we need it \citep{smadis2011}.
 
@@
     \end{tabular}
-    \label{table:coverage2}
+    \label{table:coverage}
     \caption{Evaluation metrics.}
